---
output: github_document
pandoc_args: ["+raw_html"]
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
---




<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "doc/Figures/README-",
                      out.width = "100%")
# for tibbles...
#options(pillar.neg=F, # do no print neg number in red
 #       pillar.subtle=F, # turn off highlighting of significant digits
 #       tibble.width = 170) # default=95, increase it to make it readable

```

```{r include = F}
if (!require(pacman)) install.packages("pacman")
p_load(dplyr, rio, knitr, kable, kableExtra)
```


## 📝 CDT25 Genetics Module: Summary statistics

Previously, summary statistics for many phenotypes could be extracted directly from the [IEU OpenGWAS repository](https://gwas.mrcieu.ac.uk/) using the `extract_instruments` function in the `TwoSampleMR` package. However, this function is no longer functional. Below, you will find direct links to the summary statistics for exposures and outcomes relevant to this tutorial and your projects.


### ☀️ Exposures

<span style="font-size:14px;">
**Note:** Exposure datasets include only genome-wide significant variants (*P* < 5e-8).
</span>


##### Smoking initiation (GSCAN European)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/smoking_initiation_liu_(GSCAN)_eur.csv?download=1"))`



##### Lifetime smoking index (European)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/lifetime_smoking_munafo_eur.csv?download=1"))`

##### LDL-C (Various)

For more information on these data sets, see [Table 1](#table1)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/ldl_willer.csv?download=1"))`

`data <- read.csv(url("https://zenodo.org/record/14871196/files/ldl_barton.csv?download=1"))`

`data <- read.csv(url("https://zenodo.org/record/14871196/files/ldl_wojcik.csv?download=1"))`


##### Tryglicerides (Global Lipids Genetics Consortium; Multiancestry)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/trigs_willer_(GLGC)_multi.csv?download=1"))`

##### Tryglicerides (UK Biobank)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/trigs_richardson_(UKB)_eur.csv?download=1"))`

##### Body mass index (GIANT; European)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/bmi_yengo_(GIANT)_eur.csv?download=1"))`

##### Body mass index (UK Biobank)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/bmi_neale_(UKB)_eur.csv?download=1"))`

##### Heart rate (Hrgene; European)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/heart_rate_denHoad_(HRgene)_eur.csv?download=1"))`

##### Heart rate (UK Biobank)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/heart_rate_Elsworth_(UKB)_eur.csv?download=1"))`


### ⚕️ Outcomes


##### Coronary Heart Disease (CARDIoGRAMplusC4D; European)*

`data <- read.csv(url("https://zenodo.org/record/14871196/files/chd_aragam_matched_ldl.csv?download=1"))`

<span style="font-size:14px;">
*These variants are matched to the LDL-C lists for the purpose of the practical
</span>
##### All stroke (GIGASTROKE; European)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/AS_mishra_(GIGA)_eur.csv?download=1"))`

##### All stroke (GIGASTROKE; Multiancestry)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/AS_mishra_(GIGA)_multi.csv?download=1"))`

##### All ischaemic stroke (GIGASTROKE; European)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/AIS_mishra_(GIGA)_eur.csv?download=1"))`

##### All ischaemic stroke (GIGASTROKE; Multiancestry)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/AIS_mishra_(GIGA)_multi.csv?download=1"))`

##### Cardioembolic stroke (GIGASTROKE; European)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/CES_mishra_(GIGA)_eur.csv?download=1"))`

##### Cardioembolic stroke (GIGASTROKE; Multiancestry)

`data <- read.csv(url("https://zenodo.org/record/14871196/files/CES_mishra_(GIGA)_multi.csv?download=1"))`


### ⚠️ Handling Timeout Errors for `read.csv`

If you encounter **timeout errors** when reading in large files, you can try increasing the timeout limit:

```{r, eval=FALSE}
install.packages("RCurl")
library(RCurl)
curlSetOpt(timeout = 600)
```

If that does not work, you may need to **download the file directly** and then load it into R:

```{r, eval=FALSE}
download.file("https://zenodo.org/record/14871196/files/smoking_initiation_liu_(GSCAN)_eur.csv",  
              "smoking_init.csv")
data <- read.csv("smoking_init.csv")
```

### 📊 Information on Practical data sets

<a id="table1"></a>
**Table 1: Information for exposure data and outcome data sets used in practical**


```{r, echo=FALSE, message=FALSE}
# Create the table as a data frame
table <- data.frame(
  Trait = c("LDL-C", "LDL-C", "LDL-C", "CHD"),
  Author = c("Barton", "Willer", "Wojcik", "Aragam"),
  PMID = c(34226706, 24097068, 31217584, 36474045),
  Year = c(2021, 2013, 2019, 2022),
  Pop = c("European*", "Mixed", "Hispanic/Latin American", "European"),
  Variants = c("4M", "2M", "3M", "20M"),
  `Sample size` = c("437,068", "173,082", "17,373", "1.3M (181K cases†)")
)


# Create a formatted table with column width adjustments
table %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size=11) %>%
  #column_spec(1, width = "2in") %>%  # Wider first column (PRS names)
  #column_spec(2:4, width = "1.5in") %>%  # Narrower columns for Cases/Controls, OR, P-value
  footnote(symbol = c("UK Biobank", "10% from UK Biobank"))

```




### 🙋 Questions?

Please email [Adam Von Ende](mailto:adam.vonende@ndph.ox.ac.uk) or [Elsa Valdez-Márquez](mailto:elsa.valdez-marquez@ndph.ox.ac.uk).
